---
output:
  md_document:
    variant: markdown_github
---
```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
rvers <- stringr::str_match(grep("R \\(", description, value = TRUE), "[0-9]{1,4}\\.[0-9]{1,4}\\.[0-9]{1,4}")[1,1]
version <- gsub(" ", "", gsub("Version:", "", grep("Version:", description, value = TRUE)))
```

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://tidyverse.org/lifecycle/#maturing)
[![Travis-CI Build Status](https://travis-ci.org/thierrygosselin/radiator.svg?branch=master)](https://travis-ci.org/thierrygosselin/radiator)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/thierrygosselin/radiator?branch=master&svg=true)](https://ci.appveyor.com/project/thierrygosselin/radiator)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/radiator)](http://cran.r-project.org/package=radiator)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![DOI](https://zenodo.org/badge/14548/thierrygosselin/radiator.svg)](https://zenodo.org/badge/latestdoi/14548/thierrygosselin/radiator)
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg)](commits/master)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](/commits/master)

---



# radiator: an R package for RADseq Data Exploration, Manipulation and Visualization

This is the development page of the **radiator**, 
if you want to help, see [contributions section](https://github.com/thierrygosselin/radiator#contributions)

Most genomic analysis look for patterns and trends with various statistics. 
Bias, noise and outliers can have bounded influence on estimators and interfere
with polymorphism discovery. Avoid bad data exploration and control the impact
of filters on your downstream genetic analysis. Use radiator to: import, explore,
manipulate, visualize, filter, impute and export your GBS/RADseq data.

**radiator** is designed and optimized for fast computations using Genomic Data Structure [GDS](http://zhengxwen.github.io/gdsfmt) file format and data science packages in 
[tiverse](https://www.tidyverse.org). **radiator** handles VCF files with millions of
SNPs and files of several GB.

## Installation
To try out the dev version of **radiator**, copy/paste the code below:

```r
if (!require("pak")) install.packages("pak")
pak::pkg_install("thierrygosselin/radiator")
library(radiator)
```

To minimize dependencies, just the basic required packages are installed with
the command above. If you want the full suits of functions and don't want to
be preoccupied, run:

```r
radiator::radiator_pkg_install() # that's it. It will update, when necessary, radiator.
```
[Computer setup and troubleshooting vignette](http://thierrygosselin.github.io/assigner/articles/rad_genomics_computer_setup.html)


## [Learning radiator](https://thierrygosselin.github.io/radiator/articles/getting_started.html)

## [Overview](https://thierrygosselin.github.io/radiator/articles/getting_started.html#overview)

## [New features](https://thierrygosselin.github.io/radiator/news/index.html)
Change log, version, new features and bug history lives in the [NEWS.md file](https://thierrygosselin.github.io/radiator/news/index.html)

## Life cycle

**DArT users**: 

* `filter_dart`: is now deprecated. Please use `filter_rad`.
* `tidy_dart` and `tidy_silico_dart`: are now deprecated. 
Please use `read_dart` for all the 4 DArT files recognized by radiator.

**Missing data: visualization and imputations**

Visualizing missing data and it's imputations requires special attention that fall 
outside the scope of **radiator**. 
Inside my package called [grur](https://github.com/thierrygosselin/grur), users
can **visualize patterns of missingness** associated with different variables 
(lanes, chips, sequencers, populations, sample sites, reads/samples, homozygosity, etc).
Several **Map-independent imputations** of missing genotypes are available:
**Random Forests** (on-the-fly-imputations or predictive modeling), 
**Extreme Gradient Tree Boosting**, 
Strawman imputations (~ max/mean/mode: the most frequently observed, non-missing genotypes is used).
Imputations can be conducted **overall samples** or **by populations/strata/grouping**.
`radiator::genomic_converter` is integrated with the imputation function of **grur**.

## [Vignettes](https://thierrygosselin.github.io/radiator/articles/index.html)

## Citation:
To get the citation, inside R:
```r
citation("radiator")
```
